Reanalysis Step 1: Data Preparation

### 2025-08-06 reviewer-response.Rmd

---
title: "Reviewer Response Analysis"
author: "Erik Westlund"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Load required libraries and settings
source("settings.R")
source("functions.R")

# Load all required libraries
load_required_libraries()
```

# Data Preparation

```{r load_data}
# Load and prepare data
data_path <- file.path(s_root, "Gina's Project/df_Final_240909.csv")
data <- load_and_prepare_data(data_path)

# Apply the same exclusions as main analysis
m_data <- apply_exclusions(data)

cat(sprintf("Total observations after exclusions: %d\n", nrow(m_data)))
cat(sprintf("Total unique patients: %d\n", length(unique(m_data$e_mrn_deidentified))))
```

## Load or Generate IPTW Weights

```{r twang_weights}
# Load cached results if available
cached_results <- load_cached_results("ps_results")

# Load or generate twang weights for IPTW
if("twang_gbm" %in% names(cached_results)) {
  cat("Loading cached twang model for IPTW weights...\n")
  twang_model <- cached_results$twang_gbm$result
} else {
  cat("Running twang model to generate IPTW weights...\n")
  # Get the matching formula from functions.R
  matchingFormula <- get_matching_formula()

  # Run twang ps model
  library(twang)
  twang_model <- ps(
    formula = matchingFormula, 
    data = data.frame(m_data),
    n.trees = 10000,
    interaction.depth = 3,
    shrinkage = 0.01,
    estimand = "ATT",
    stop.method = c("es.mean", "ks.max")
  )

  # Save for future use
  if(!dir.exists("ps_results")) {
    dir.create("ps_results")
  }
  saveRDS(list(result = twang_model), file = "ps_results/twang_gbm.rds")
}

# Extract IPTW weights from twang model
m_data$twang_att_w <- get.weights(twang_model, stop.method = "es.mean")

# Summary of weights
summary(m_data$twang_att_w)
```

# Reviewer Question 1: Subgroup Testing with Formal Interaction Terms

**Reviewer Comment:** "Inadequate Subgroup Testing: Visual forest‑plot overlap doesn't prove consistent effects across subgroups (e.g., PDR+anti‑VEGF vs. PDR+PRP). Without formal interaction testing, subgroup claims lack statistical backing."

**Response:** We perform formal interaction testing using two models:
1. Interaction between lapse status and DR severity
2. Interaction between lapse status and treatment type

## Create Analysis Variables

```{r create_variables}
# Create the treatment type variable with expanded categories
m_data <- m_data |>
  mutate(
    treatment_type = case_when(
      anti_VEGF == 1 ~ "anti_VEGF",
      PRP_flag == 1 ~ "PRP",
      other_inject == 1 ~ "other_inject",
      focal_laser_flag == 1 ~ "focal_laser",
      TRUE ~ "no_treatment"
    ),
    treatment_type = factor(treatment_type, 
                           levels = c("no_treatment", "anti_VEGF", "PRP", 
                                    "other_inject", "focal_laser")),
    # Also create a severe DR indicator
    severe_dr = (NPDR == "Severe" | PDR == "Present")
  )

# Check distribution of treatment types
table(m_data$treatment_type)

# Check severe DR distribution
table(m_data$severe_dr)
```

## Define Model Formulas

```{r define_formulas}
# Define the base formula (same as analysisFormulaFull but without the outcome)
base_covariates <- c(
  "gender_cat",
  "race_ethnic_cat", 
  "insurance_cat",
  "age_cat",
  "CCI",
  "DCSI",
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat",
  "catsurg_before_hitplus_cat"
)

# Model 1: Interaction with DR severity (person_dr)
# person_dr: 0 = No DR, 1 = NPDR, 2 = PDR
interaction_dr_formula <- as.formula(paste(
  "outcome_va_vi_binary ~ ever_lapse_binary * factor(person_dr) +",
  paste(base_covariates, collapse = " + ")
))

# Model 2: Interaction with treatment type
interaction_treatment_formula <- as.formula(paste(
  "outcome_va_vi_binary ~ ever_lapse_binary * treatment_type +",
  paste(base_covariates, collapse = " + ")
))

print("DR Interaction Formula:")
print(interaction_dr_formula)

print("\nTreatment Interaction Formula:")
print(interaction_treatment_formula)
```

## Run Interaction Models

```{r interaction_models, results='asis'}
# Store results for both weighted and unweighted analyses
interaction_results_list <- list()

# Create separate data subsets for analysis
data_subsets <- list(
  all_cases = m_data,
  severe_dr_only = m_data |> filter(severe_dr == TRUE)
)

# Loop through both data subsets and both weighted and unweighted designs
for(subset_name in names(data_subsets)) {

  current_data <- data_subsets[[subset_name]]

  cat("\n\n## Data Subset:", toupper(subset_name), "\n")
  cat("Sample size:", nrow(current_data), "\n\n")

  # Update weights for current subset
  if("twang_att_w" %in% names(current_data)) {
    # Weights already exist from full dataset
  } else {
    current_data$twang_att_w <- m_data$twang_att_w[m_data$e_mrn_deidentified %in% current_data$e_mrn_deidentified]
  }

  for(weight_type in c("unweighted", "weighted")) {

    cat("\n### Analysis Type:", toupper(weight_type), "\n\n")

    # Create survey design for current subset
    current_design <- if(weight_type == "unweighted") {
      svydesign(
        ids = ~e_mrn_deidentified,
        data = current_data,
        weights = ~1
      )
    } else {
      svydesign(
        ids = ~e_mrn_deidentified,
        data = current_data,
        weights = ~twang_att_w
      )
    }

    # Fit Model 1: DR severity interaction
    cat("\n#### DR Severity Interaction Model\n\n")
    model_dr_interaction <- svyglm(
      formula = interaction_dr_formula,
      design = current_design,
      family = quasibinomial()
    )

    # Summary of DR interaction model
    summary_dr <- summary(model_dr_interaction)

    # Extract interaction p-values for DR model
    dr_interaction_terms <- grep("ever_lapse_binary:factor\\(person_dr\\)", 
                                names(coef(model_dr_interaction)), value = TRUE)
    dr_interaction_pvals <- if(length(dr_interaction_terms) > 0) {
      summary_dr$coefficients[dr_interaction_terms, "Pr(>|t|)"]
    } else {
      numeric(0)
    }

    cat("\n**DR Severity Interaction P-values:**\n\n")
    if(length(dr_interaction_terms) > 0) {
      dr_results <- data.frame(
        Term = dr_interaction_terms,
        P_Value = round(dr_interaction_pvals, 4),
        Significant = ifelse(dr_interaction_pvals < 0.05, "Yes", "No")
      )
      print(knitr::kable(dr_results))
    } else {
      cat("No interaction terms found\n")
    }

    # Fit Model 2: Treatment type interaction  
    cat("\n#### Treatment Type Interaction Model\n\n")
    model_treatment_interaction <- svyglm(
      formula = interaction_treatment_formula,
      design = current_design,
      family = quasibinomial()
    )

    # Summary of treatment interaction model
    summary_treatment <- summary(model_treatment_interaction)

    # Extract interaction p-values for treatment model
    treatment_interaction_terms <- grep("ever_lapse_binary:treatment_type", 
                                       names(coef(model_treatment_interaction)), value = TRUE)
    treatment_interaction_pvals <- if(length(treatment_interaction_terms) > 0) {
      summary_treatment$coefficients[treatment_interaction_terms, "Pr(>|t|)"]
    } else {
      numeric(0)
    }

    cat("\n**Treatment Type Interaction P-values:**\n\n")
    if(length(treatment_interaction_terms) > 0) {
      tx_results <- data.frame(
        Term = treatment_interaction_terms,
        P_Value = round(treatment_interaction_pvals, 4),
        Significant = ifelse(treatment_interaction_pvals < 0.05, "Yes", "No")
      )
      print(knitr::kable(tx_results))
    } else {
      cat("No interaction terms found\n")
    }

    # Store results for this weight type and subset
    if(!subset_name %in% names(interaction_results_list)) {
      interaction_results_list[[subset_name]] <- list()
    }
    interaction_results_list[[subset_name]][[weight_type]] <- list(
      dr_model = model_dr_interaction,
      treatment_model = model_treatment_interaction,
      dr_pvalues = dr_interaction_pvals,
      treatment_pvalues = treatment_interaction_pvals
    )
  } # end weight_type loop
} # end subset_name loop
```

## Summary of Interaction Tests

```{r summary_table}
# Create a comparison summary table of interaction effects
comparison_rows <- list()
for(subset_name in names(interaction_results_list)) {
  for(weight_type in c("unweighted", "weighted")) {
    subset_results <- interaction_results_list[[subset_name]][[weight_type]]

    comparison_rows[[paste(subset_name, weight_type, sep="_")]] <- data.frame(
      Subset = subset_name,
      Model = c("DR Severity", "Treatment Type"),
      Weight_Type = weight_type,
      N_Interactions = c(
        length(subset_results$dr_pvalues),
        length(subset_results$treatment_pvalues)
      ),
      Min_P_Value = c(
        ifelse(length(subset_results$dr_pvalues) > 0, 
               min(subset_results$dr_pvalues), NA),
        ifelse(length(subset_results$treatment_pvalues) > 0, 
               min(subset_results$treatment_pvalues), NA)
      ),
      Significant_at_05 = c(
        sum(subset_results$dr_pvalues < 0.05),
        sum(subset_results$treatment_pvalues < 0.05)
      )
    )
  }
}
comparison_results <- do.call(rbind, comparison_rows)

knitr::kable(comparison_results, 
             caption = "Summary of Interaction Test Results",
             digits = 4)
```

## Interpretation of Results

```{r interpretation}
# Check for significant interactions and report
for(subset_name in names(interaction_results_list)) {
  cat("\n### ", toupper(subset_name), "\n")

  # Check unweighted results
  unweighted_results <- interaction_results_list[[subset_name]][["unweighted"]]
  if(any(unweighted_results$dr_pvalues < 0.05)) {
    cat("\n**UNWEIGHTED: Significant DR severity interactions detected**\n")
    sig_terms <- names(unweighted_results$dr_pvalues)[
      unweighted_results$dr_pvalues < 0.05
    ]
    cat("Significant terms:", paste(sig_terms, collapse = ", "), "\n")
  } else {
    cat("\n**UNWEIGHTED: No significant DR severity interactions**\n")
  }

  if(any(unweighted_results$treatment_pvalues < 0.05)) {
    cat("\n**UNWEIGHTED: Significant treatment type interactions detected**\n")
    sig_terms <- names(unweighted_results$treatment_pvalues)[
      unweighted_results$treatment_pvalues < 0.05
    ]
    cat("Significant terms:", paste(sig_terms, collapse = ", "), "\n")
  } else {
    cat("\n**UNWEIGHTED: No significant treatment type interactions**\n")
  }

  # Check weighted results
  weighted_results <- interaction_results_list[[subset_name]][["weighted"]]
  if(any(weighted_results$dr_pvalues < 0.05)) {
    cat("\n**WEIGHTED (IPTW): Significant DR severity interactions detected**\n")
    sig_terms <- names(weighted_results$dr_pvalues)[
      weighted_results$dr_pvalues < 0.05
    ]
    cat("Significant terms:", paste(sig_terms, collapse = ", "), "\n")
  } else {
    cat("\n**WEIGHTED (IPTW): No significant DR severity interactions**\n")
  }

  if(any(weighted_results$treatment_pvalues < 0.05)) {
    cat("\n**WEIGHTED (IPTW): Significant treatment type interactions detected**\n")
    sig_terms <- names(weighted_results$treatment_pvalues)[
      weighted_results$treatment_pvalues < 0.05
    ]
    cat("Significant terms:", paste(sig_terms, collapse = ", "), "\n")
  } else {
    cat("\n**WEIGHTED (IPTW): No significant treatment type interactions**\n")
  }
}

cat("\n\n**NOTE:** Interaction p-values test whether the effect of lapsing differs")
cat("\nsignificantly across subgroups. Non-significant p-values suggest")
cat("\nconsistent effects across groups, supporting the main analysis.")
cat("\nThe IPTW-weighted analysis adjusts for confounding using propensity scores.\n")
```

# Visualization of Interaction Effects

```{r visualization_functions}
# Function to extract predicted probabilities for interaction plots
extract_predictions <- function(model, data, interaction_var, fixed_vars = NULL) {
  # Create prediction data
  pred_data <- expand.grid(
    ever_lapse_binary = c(0, 1),
    stringsAsFactors = FALSE
  )

  # Add the interaction variable levels
  if(interaction_var == "person_dr") {
    pred_data <- expand.grid(
      ever_lapse_binary = c(0, 1),
      person_dr = unique(data$person_dr),
      stringsAsFactors = FALSE
    )
    pred_data$person_dr <- factor(pred_data$person_dr)
  } else if(interaction_var == "treatment_type") {
    pred_data <- expand.grid(
      ever_lapse_binary = c(0, 1),
      treatment_type = levels(data$treatment_type),
      stringsAsFactors = FALSE
    )
  }

  # Set other variables to their reference levels or means
  for(var in all.vars(formula(model))[-1]) {
    if(!var %in% names(pred_data)) {
      if(is.factor(data[[var]]) || is.character(data[[var]])) {
        pred_data[[var]] <- factor(levels(factor(data[[var]]))[1], 
                                  levels = levels(factor(data[[var]])))
      } else {
        pred_data[[var]] <- mean(data[[var]], na.rm = TRUE)
      }
    }
  }

  # Get predictions
  pred_data$predicted <- predict(model, newdata = pred_data, type = "response")
  pred_data$se <- predict(model, newdata = pred_data, type = "response", se.fit = TRUE)$se.fit
  pred_data$lower <- pred_data$predicted - 1.96 * pred_data$se
  pred_data$upper <- pred_data$predicted + 1.96 * pred_data$se

  return(pred_data)
}
```

## DR Severity Interaction Plots

```{r dr_interaction_plots, fig.height=8}
library(ggplot2)
library(dplyr)
library(tidyr)

# Create plots for each subset and weight combination
plot_list <- list()

for(subset_name in names(interaction_results_list)) {
  for(weight_type in c("unweighted", "weighted")) {

    current_results <- interaction_results_list[[subset_name]][[weight_type]]
    current_data <- data_subsets[[subset_name]]

    # Plot 1: DR Severity Interaction
    dr_pred <- extract_predictions(current_results$dr_model, current_data, "person_dr")
    dr_pred$lapse_status <- factor(dr_pred$ever_lapse_binary, 
                                   levels = c(0, 1),
                                   labels = c("No Lapse", "Lapse"))
    dr_pred$dr_label <- factor(dr_pred$person_dr,
                               levels = c(0, 1, 2),
                               labels = c("No DR", "NPDR", "PDR"))

    p_dr <- ggplot(dr_pred, aes(x = dr_label, y = predicted, 
                                color = lapse_status, group = lapse_status)) +
      geom_point(position = position_dodge(0.2), size = 3) +
      geom_errorbar(aes(ymin = lower, ymax = upper), 
                   width = 0.2, position = position_dodge(0.2)) +
      geom_line(position = position_dodge(0.2), alpha = 0.5) +
      labs(title = paste("DR Severity Interaction:", subset_name, "-", weight_type),
           x = "DR Severity", 
           y = "Predicted Probability of Vision Impairment",
           color = "Lapse Status") +
      theme_minimal() +
      theme(legend.position = "bottom") +
      scale_y_continuous(limits = c(0, 1))

    print(p_dr)

    # Store plot
    plot_name_dr <- paste(subset_name, weight_type, "dr", sep = "_")
    plot_list[[plot_name_dr]] <- p_dr
  }
}
```

## Treatment Type Interaction Plots

```{r treatment_interaction_plots, fig.height=8}
for(subset_name in names(interaction_results_list)) {
  for(weight_type in c("unweighted", "weighted")) {

    current_results <- interaction_results_list[[subset_name]][[weight_type]]
    current_data <- data_subsets[[subset_name]]

    # Plot 2: Treatment Type Interaction
    tx_pred <- extract_predictions(current_results$treatment_model, current_data, "treatment_type")
    tx_pred$lapse_status <- factor(tx_pred$ever_lapse_binary,
                                   levels = c(0, 1),
                                   labels = c("No Lapse", "Lapse"))

    p_tx <- ggplot(tx_pred, aes(x = treatment_type, y = predicted,
                                color = lapse_status, group = lapse_status)) +
      geom_point(position = position_dodge(0.2), size = 3) +
      geom_errorbar(aes(ymin = lower, ymax = upper),
                   width = 0.2, position = position_dodge(0.2)) +
      geom_line(position = position_dodge(0.2), alpha = 0.5) +
      labs(title = paste("Treatment Type Interaction:", subset_name, "-", weight_type),
           x = "Treatment Type",
           y = "Predicted Probability of Vision Impairment",
           color = "Lapse Status") +
      theme_minimal() +
      theme(legend.position = "bottom",
            axis.text.x = element_text(angle = 45, hjust = 1)) +
      scale_y_continuous(limits = c(0, 1))

    print(p_tx)

    # Store plot
    plot_name_tx <- paste(subset_name, weight_type, "treatment", sep = "_")
    plot_list[[plot_name_tx]] <- p_tx
  }
}
```

## Forest Plot of Treatment-Specific Effects

```{r forest_plot, fig.width=12, fig.height=10}
# Extract coefficients for lapse effect in each treatment subgroup
treatment_effects <- data.frame()

for(subset_name in names(interaction_results_list)) {
  for(weight_type in c("unweighted", "weighted")) {
    model <- interaction_results_list[[subset_name]][[weight_type]]$treatment_model
    coef_summary <- summary(model)$coefficients

    # Get main effect
    main_effect <- coef_summary["ever_lapse_binary", "Estimate"]
    main_se <- coef_summary["ever_lapse_binary", "Std. Error"]

    # Get interaction effects
    interaction_terms <- grep("ever_lapse_binary:treatment_type", 
                            rownames(coef_summary), value = TRUE)

### 2025-08-14 reanalysis-1-data_preparation.qmd
update

Loading the new combined dataset for reanalysis with multiple imputation support.

```{r load-data}
# Load the combined dataset
cat("Loading combined dataset from:", reanalysis_data_file_path, "\n")
df_combined <- readr::read_csv(reanalysis_data_file_path)

# Display basic information about the dataset
cat("Dataset dimensions:", nrow(df_combined), "rows x", ncol(df_combined), "columns\n")
glimpse(df_combined)
```

## Data Structure Exploration

```{r data-structure}
# Summary of the dataset
cat("=== DATASET SUMMARY ===\n")
summary(df_combined)

# Check for key variables that we expect
expected_vars <- c(
  "e_mrn_deidentified",
  "person_ever_lapse", 
  "outcome_VA_logMAR",
  "age", "gender", "race_ethnicity_gp", "insurance_gp",
  "CCI", "DCSI",
  "No_DR", "NPDR", "PDR",
  "anti_VEGF", "PRP_flag", "other_inject", "focal_laser_flag"
)

cat("\n=== KEY VARIABLE CHECK ===\n")
for(var in expected_vars) {
  if(var %in% names(df_combined)) {
    cat("✓", var, "- present\n")
  } else {
    cat("✗", var, "- MISSING\n")
  }
}

# Display first few rows
cat("\n=== FIRST 10 ROWS ===\n")
head(df_combined, 10)
```

@@

```{r missing-data}
# Overall missing data pattern
cat("=== MISSING DATA SUMMARY ===\n")
missing_summary <- df_combined |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |>
  mutate(
    missing_percent = round(missing_count / nrow(df_combined) * 100, 2)
  ) |>
  arrange(desc(missing_count))

print(missing_summary, n = 20)

# Visualize missing data pattern using VIM package
if(require(VIM, quietly = TRUE)) {
  cat("\n=== MISSING DATA PATTERN VISUALIZATION ===\n")
  aggr(df_combined, col = c('navyblue','red'), 
       numbers = TRUE, sortVars = TRUE, labels = names(df_combined), 
       cex.axis = 0.7, gap = 3, 
       ylab = c("Histogram of missing data","Pattern"))
} else {
  cat("VIM package not available for missing data visualization\n")
}
```

```{r variable-creation}
# Apply the existing data preparation functions
cat("=== APPLYING DATA TRANSFORMATIONS ===\n")

# Check if we can use the existing load_and_prepare_data function
# or if we need to adapt it for the new dataset
tryCatch({
  # Test if the existing function works with the new data structure
  # We'll create a temporary file path for testing
  temp_path <- tempfile(fileext = ".csv")
  write_csv(df_combined, temp_path)

  # Try using existing function
  df_prepared <- load_and_prepare_data(temp_path)

  cat("✓ Successfully applied existing data preparation functions\n")
  cat("Prepared dataset dimensions:", nrow(df_prepared), "rows x", ncol(df_prepared), "columns\n")

  # Clean up temp file
  unlink(temp_path)

}, error = function(e) {
  cat("✗ Error applying existing functions:", e$message, "\n")
  cat("Will need to adapt functions for new dataset structure\n")

  # For now, create a basic preparation
  df_prepared <- df_combined |>
    mutate(
      # Create basic outcome variables
      outcome_va_vi = if("outcome_VA_logMAR" %in% names(df_combined)) {
        outcome_VA_logMAR >= 0.3
      } else {
        NA
      },
      outcome_va_vi_binary = if("outcome_VA_logMAR" %in% names(df_combined)) {
        as.numeric(outcome_VA_logMAR >= 0.3)
      } else {
        NA
      },

      # Create lapse variable
      ever_lapse_binary = if("person_ever_lapse" %in% names(df_combined)) {
        as.numeric(person_ever_lapse == TRUE)
      } else {
        NA
      }
    )

  cat("Applied basic variable creation as fallback\n")
})

# Display the prepared dataset structure
cat("\n=== PREPARED DATASET STRUCTURE ===\n")
glimpse(df_prepared)
```

## Data Quality Checks

```{r data-quality}
# Check key variables for analysis
cat("=== DATA QUALITY ASSESSMENT ===\n")

# Check outcome variable distribution
if("outcome_va_vi_binary" %in% names(df_prepared)) {
  cat("Outcome variable (vision impairment) distribution:\n")
  table(df_prepared$outcome_va_vi_binary, useNA = "ifany")
  cat("\n")
}

# Check exposure variable distribution  
if("ever_lapse_binary" %in% names(df_prepared)) {
  cat("Exposure variable (ever lapse) distribution:\n")
  table(df_prepared$ever_lapse_binary, useNA = "ifany")
  cat("\n")
}

# Check for potential exclusions
cat("=== POTENTIAL EXCLUSIONS ===\n")

exclusion_checks <- list(
  "Missing outcome" = sum(is.na(df_prepared$outcome_va_vi_binary)),
  "Missing exposure" = sum(is.na(df_prepared$ever_lapse_binary)),
  "Missing CCI" = if("CCI" %in% names(df_prepared)) sum(is.na(df_prepared$CCI)) else "Variable not found",
  "Missing DCSI" = if("DCSI" %in% names(df_prepared)) sum(is.na(df_prepared$DCSI)) else "Variable not found"
)

for(check_name in names(exclusion_checks)) {
  cat(check_name, ":", exclusion_checks[[check_name]], "\n")
}
```

## Comparison with Original Dataset

```{r compare-original}
# Load original dataset for comparison if available
if(file.exists(original_analysis_file_path)) {
  cat("=== COMPARISON WITH ORIGINAL DATASET ===\n")

  df_original <- readr::read_csv(original_analysis_file_path)

  cat("Original dataset dimensions:", nrow(df_original), "rows x", ncol(df_original), "columns\n")
  cat("Combined dataset dimensions:", nrow(df_combined), "rows x", ncol(df_combined), "columns\n")

  # Compare key variables
  common_vars <- intersect(names(df_original), names(df_combined))
  cat("Common variables:", length(common_vars), "\n")
  cat("Variables only in original:", length(setdiff(names(df_original), names(df_combined))), "\n")
  cat("Variables only in combined:", length(setdiff(names(df_combined), names(df_original))), "\n")

  # Show some of the different variables
  cat("\nVariables only in original (first 10):\n")
  cat(paste(head(setdiff(names(df_original), names(df_combined)), 10), collapse = ", "), "\n")

  cat("\nVariables only in combined (first 10):\n") 
  cat(paste(head(setdiff(names(df_combined), names(df_original)), 10), collapse = ", "), "\n")

} else {
  cat("Original analysis file not found at:", original_analysis_file_path, "\n")
  cat("Cannot perform comparison\n")
}
```

## Summary and Next Steps

```{r summary}
cat("=== DATA PREPARATION SUMMARY ===\n")
cat("✓ Successfully loaded combined dataset\n")
cat("✓ Assessed missing data patterns\n")
cat("✓ Applied data transformations\n")
cat("✓ Performed quality checks\n")

if(exists("df_prepared")) {
  cat("✓ Final prepared dataset ready with", nrow(df_prepared), "observations\n")

  # Save prepared dataset for next steps
  prepared_data_path <- "prepared_data.rds"
  saveRDS(df_prepared, prepared_data_path)
  cat("✓ Saved prepared dataset to:", prepared_data_path, "\n")
} else {
  cat("✗ Issue with data preparation - needs attention\n")
}

cat("\n=== NEXT STEPS ===\n")
cat("1. Review column structure compatibility with functions.R\n")
cat("2. Implement multiple imputation strategy\n")
cat("3. Apply propensity score methods to imputed datasets\n")
cat("4. Pool results across imputations\n")
```

## Session Info

```{r session-info}
sessionInfo()
```

>>>>

```{r load-data}
# Load the combined dataset
df_combined <- readr::read_csv(reanalysis_data_file_path)
```

## Define Analysis Variables

```{r define-variables}
# Variables for propensity score model
ps_vars <- c(
  "baseline_VA_logMAR",
  "gender",
  "race_ethnicity_gp", 
  "insurance_gp",
  "age",
  "CCI",
  "DCSI",
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  "Glaucoma_bef_hitplus",
  "Otherretina_bef_hitplus",
  "cataract_surgery_bef_hitplus"
)

# Variables for outcome models
outcome_vars <- c(
  "outcome_VA_logMAR",
  "person_ever_lapse"
)

# Variables for DR severity and treatment categorization
dr_vars <- c(
  "No_DR",
  "NPDR",
  "PDR"
)

# All analysis variables
all_analysis_vars <- c(
  "e_mrn_deidentified",  # Patient ID
  ps_vars,
  outcome_vars,
  dr_vars
)
```

```{r create-variables}
# Create analysis dataset with necessary transformations
df_analysis <- df_combined |>
  mutate(
    # Outcome variables
    outcome_va_vi = outcome_VA_logMAR >= 0.3,
    outcome_va_vi_binary = if_else(outcome_va_vi, 1, 0),
    outcome_va_blind = outcome_VA_logMAR >= 1,
    outcome_va_blind_binary = if_else(outcome_va_blind, 1, 0),

    # Exposure variable
    ever_lapse_binary = if_else(person_ever_lapse == TRUE, 1, 0),

    # Demographic factors
    gender_cat = factor(gender),
    race_ethnic_cat = factor(race_ethnicity_gp),
    insurance_cat = factor(insurance_gp),
    age_cat = cut(age, breaks = c(0, 20, 45, 65, Inf), 
                  labels = c("0-20", "21-45", "46-65", "65+")),

    # DR severity (0 = No DR, 1 = NPDR, 2 = PDR)
    person_dr = case_when(
      No_DR == TRUE ~ 0,
      NPDR == "Mild" | NPDR == "Moderate" | NPDR == "Severe" ~ 1,
      PDR == "Present" ~ 2
    ),

    # DR severity detailed (for stratified analyses)
    person_dr_severity = case_when(
      No_DR == TRUE ~ 0,
      NPDR == "Mild" ~ 1,
      NPDR == "Moderate" ~ 2,
      NPDR == "Severe" ~ 3,
      PDR == "Present" ~ 4
    ),

    # DR severity factor for interaction models
    dr_severity = factor(person_dr,
                        levels = c(0, 1, 2),
                        labels = c("No_DR", "NPDR", "PDR")),

    # Treatment type categorization
    treatment_type = case_when(
      anti_VEGF == 1 ~ "anti_VEGF",
      PRP_flag == 1 ~ "PRP",
      other_inject == 1 | focal_laser_flag == 1 ~ "other_treatment",
      TRUE ~ "no_treatment"
    ),
    treatment_type = factor(treatment_type, 
                           levels = c("no_treatment", "anti_VEGF", "PRP", "other_treatment")),

    # Binary treatment indicator
    any_treatment = as.numeric(anti_VEGF == 1 | PRP_flag == 1 | 
                              other_inject == 1 | focal_laser_flag == 1),

    # Comorbidity indicators as factors
    glaucoma_bef_hitplus_cat = factor(if_else(Glaucoma_bef_hitplus == "Present", 1, 0)),
    otherretina_bef_hitplus_cat = factor(if_else(Otherretina_bef_hitplus == "Present", 1, 0)),
    catsurg_before_hitplus_cat = factor(cataract_surgery_bef_hitplus),

    # Treatment indicators as factors
    other_inject = factor(other_inject),
    anti_VEGF = factor(anti_VEGF),
    focal_laser_flag = factor(focal_laser_flag),
    PRP_flag = factor(PRP_flag)
  )
```

@@

```{r missing-assessment}
# Variables used in propensity score model
ps_model_vars <- c(
  "ever_lapse_binary",
  "baseline_VA_logMAR",
  "gender_cat",
  "race_ethnic_cat", 
  "insurance_cat",
  "age_cat",
  "CCI",
  "DCSI",
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat",
  "catsurg_before_hitplus_cat"
)

# Variables used in outcome model
outcome_model_vars <- c(
  "outcome_va_vi_binary",
  ps_model_vars  # All PS vars are also in outcome model
)

# Assess missing data for analysis variables
missing_summary <- df_analysis |>
  select(all_of(c(ps_model_vars, "outcome_va_vi_binary"))) |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  mutate(
    pct_missing = round(n_missing / nrow(df_analysis) * 100, 2),
    var_type = case_when(
      variable %in% c("outcome_va_vi_binary") ~ "Outcome",
      variable == "ever_lapse_binary" ~ "Exposure",
      TRUE ~ "Covariate"
    )
  ) |>
  arrange(desc(n_missing))

# Display missing data summary
missing_summary |> 
  filter(n_missing > 0) |>
  kable(caption = "Variables with missing data (for imputation)")
```

## Data Summary for Analysis Variables

```{r data-summary}
# Summary of key variables
summary_vars <- c(
  "outcome_va_vi_binary",
  "ever_lapse_binary", 
  "dr_severity",
  "treatment_type",
  "age",
  "CCI",
  "DCSI"
)

# Create summary table
df_analysis |>
  select(all_of(summary_vars)) |>
  summary()

# Distribution of outcome by exposure
table(df_analysis$outcome_va_vi_binary, df_analysis$ever_lapse_binary, 
      dnn = c("Vision Impairment", "Ever Lapsed"))

# Distribution of DR severity
table(df_analysis$dr_severity, useNA = "ifany")

# Distribution of treatment types
table(df_analysis$treatment_type, useNA = "ifany")
```

## Complete Case Analysis Dataset

```{r complete-cases}
# For comparison, create complete case dataset (no missing values)
df_complete <- df_analysis |>
  filter(!is.na(CCI) & !is.na(DCSI) & !is.na(baseline_VA_logMAR))

# Summary of exclusions
n_total <- nrow(df_analysis)
n_complete <- nrow(df_complete)
n_excluded <- n_total - n_complete

tibble(
  Dataset = c("Full dataset", "Complete cases", "Excluded"),
  N = c(n_total, n_complete, n_excluded),
  Percent = c(100, round(n_complete/n_total*100, 1), round(n_excluded/n_total*100, 1))
) |>
  kable(caption = "Complete case analysis summary")
```

## Save Prepared Data

```{r save-data}
# Create directories if they don't exist
if (!dir.exists(reanalysis_data_dir)) {
  dir.create(reanalysis_data_dir, recursive = TRUE)
}

# Save the prepared dataset to network drive
saveRDS(df_analysis, file.path(reanalysis_data_dir, "df_analysis_prepared.rds"))
saveRDS(df_complete, file.path(reanalysis_data_dir, "df_complete_cases.rds"))

# Save variable lists for use in other scripts
variable_lists <- list(
  ps_model_vars = ps_model_vars,
  outcome_model_vars = outcome_model_vars,
  dr_vars = dr_vars,
  all_analysis_vars = all_analysis_vars
)
saveRDS(variable_lists, file.path(reanalysis_data_dir, "variable_lists.rds"))

# Confirm files saved
message("Data saved to: ", reanalysis_data_dir)
```

## Summary

```{r summary}
# Final summary
tibble(
  Metric = c(
    "Total observations",
    "Complete cases for main analysis", 
    "Missing CCI/DCSI",
    "Missing baseline VA",
    "Patients who lapsed",
    "Patients with vision impairment"
  ),
  N = c(
    nrow(df_analysis),
    nrow(df_complete),
    sum(is.na(df_analysis$CCI) | is.na(df_analysis$DCSI)),
    sum(is.na(df_analysis$baseline_VA_logMAR)),
    sum(df_analysis$ever_lapse_binary == 1, na.rm = TRUE),
    sum(df_analysis$outcome_va_vi_binary == 1, na.rm = TRUE)
  )
) |>
  kable(caption = "Dataset summary for reanalysis")
```

## Next Steps

1. **Multiple Imputation** (`reanalysis-2-multiple_imputation.qmd`): Impute missing values in covariates
2. **Propensity Score Analysis** (`reanalysis-3-propensity_scores.qmd`): Calculate PS using various methods on imputed datasets
3. **Outcome Analysis** (`reanalysis-4-outcome_analysis.qmd`): Fit outcome models and pool results
4. **Sensitivity Analysis** (`reanalysis-5-sensitivity.qmd`): Compare complete case vs imputed results
